Future challenges for Research Software Engineering

### 2025-05-12 future_challenges_for_RSEng.tex
Technical fixes

\subsection{Definition and Evolution of RSEs}
\begin{whatis}{What do we want to have achieved}{ethics}
The term RSE is established in the minds of university boards, as a valuable member in the academic framework.
RSEs are developed using specialized master's program, and their support is valued by researchers. We still teach researchers basic software skills, but the content of this lecture has changed drastically.
\end{whatis}
% Here we consider how the topics of RSEng will change in the face of the digitalization of society.
% Also here: What will our topics be after version control doesn't need to be taught anymore?
% Digitalisation will also feature more heavily in the domain curricula.
In this discussions the workshop participants found that
the digital skills of domain scientists will increase,
and software will become easier to interact with,
reducing reliance on RSEs to train newcomers on the most basic tasks
(JN: for example, versioning and archiving?)
(GC: Higher-level and more intuitive workflows that require less training).
User training sessions might focus on more advanced topics
that cater to a smaller audience, or involve mentoring more frequently.

@@

\subsection{Development of the RSE community}
\begin{whatis}{What do we want to have achieved}{ethics}
We have more national RSE communities that are interacting with other partners in their nation.
There are regular international conferences. The installation of RSE Master's programmes led to
increased numbers of the members of the national societies.
\end{whatis}
In recent years, we saw the formation of RSE institutes and advocacy groups, such as
the UK Software Sustainability Institute\footnote{\url{https://www.software.ac.uk}},
the US Research Software Sustainability Institute\footnote{\url{https://urssi.us}},
the European Virtual Institute for Research Software Excellence\footnote{\url{https://everse.software}},
and the EOSC Opportunity Area Expert Group 7 on Research Software%
\footnote{\url{https://eosc.eu/opportunity-area-exp/oa7-research-software}}.
We expect these initiatives to help increase the visibility and accelerate
the professionalisation%
\footnote{MM: do we want to decide for UK or US English? Or keep it mixed? Not being a native English speaker I don't know how it reads.}
of the RSE role.
In addition, a multitude of RSE communities self-organised at the national
and regional level to support more viable RSE career paths in academia.
We can expect RSEs to become more easily accessible,
possibly through centrally-funded long-term positions at research institutions.
The community of RSE's will get a significant boost in their acceptance if RSE Masters programmes
become available at various universities.

@@

\subsection{Impact of AI and automation}
\begin{whatis}{What do we want to have achieved}{ai}
RSEs are educated about AI models, and can help researchers reap their benefits, while
clearly knowing their limits.
\end{whatis}
AI is already transforming research processes by enabling rapid analysis of massive datasets, uncovering patterns and insights that would be impossible or prohibitively time-consuming for humans alone to find.
In research and development, AI accelerates scientific breakthroughs in fields like healthcare, climate science, and materials engineering by automating complex modeling and predictive tasks.
For software development, AI-powered tools automate code generation, bug detection, and testing, significantly increasing productivity and reducing human error.
Developers are shifting from manual coding to orchestrating AI-driven workflows, focusing more on high-level design, integration, and creative problem-solving while AI handles repetitive or routine tasks.
The integration of AI is democratizing access to advanced technologies and analytics, enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
As AI continues to evolve, it will further blur the lines between human and machine roles, making research and software development faster, more efficient, and more innovative than ever before.
With the general availability of AI the interaction with researchers will change at a fundamental level.
Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to it,
It can be foreseen, that AI drives this low-key interaction to a highly specialized level of knowledge,
since AI copilots are available 24/7.

@@

The ethics and values of the Research Software Engineer are grounded in their practice.
They commit to the health, safety and welfare of the public and act in the interest of society, their employer and their clients \cite{Goth2023}.

@@

\begin{whatis}{What do we want to have achieved}{academic}
School curricula have embraced the digital transformation, and have included basics of computer interaction
in every class level.
In universities, the first generations of RSE-Master graduates find their way into research departments,
and accelerate the pace of science.
\end{whatis}

@@

\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
\begin{whatis}{What do we want to have achieved}{academic}
We would like to see the emergence of more dedicated personel, i.e. other RTPs next to the RSE.
\end{whatis}
RSEs can either be embedded in research groups (belonging to the mid-level university staff),
or be members of larger RSE units. (TODO: Cite RSE groups paper)
The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.
This, however, comes with additional funding challenges,
as long-term funding is more difficult to acquire in lower levels of an organization.
Third-party funding can mitigate this issue in the mid-term.
However, the participants considered central RSE groups more likely in the future,
due to the fact that there are already examples and activities from various universities in this direction
(examples include the universities of Jena, Heidelberg, Göttingen).
We expect funding in central groups to evolve from initially temporary positions to permanent positions,
as the groups themselves prove themselves valuable.
Embedded RSEs and central groups will need to network with each other,
and potentially request and coordinate support from external RSEs.

>>>>

\subsection{Definition and Evolution of RSEs}
\begin{whatis}{What do we want to have achieved}{definition-evolution-rse}
The term RSE is established in the minds of university boards, as a valuable member in the academic framework.
RSEs are developed using specialized master's program, and their support is valued by researchers. We still teach researchers basic software skills, but the content of this lecture has changed drastically.
\end{whatis}
% Here we consider how the topics of RSEng will change in the face of the digitalization of society.
% Also here: What will our topics be after version control doesn't need to be taught anymore?
% Digitalisation will also feature more heavily in the domain curricula.
In this discussions the workshop participants found that
the digital skills of domain scientists will increase,
and software will become easier to interact with,
reducing reliance on RSEs to train newcomers on the most basic tasks
(JN: for example, versioning and archiving?)
(GC: Higher-level and more intuitive workflows that require less training).
User training sessions might focus on more advanced topics
that cater to a smaller audience, or involve mentoring more frequently.

@@

\subsection{Development of the RSE community}
\begin{whatis}{What do we want to have achieved}{development-rse-community}
We have more national RSE communities that are interacting with other partners in their nation.
There are regular international conferences. The installation of RSE Master's programmes led to
increased numbers of the members of the national societies.
\end{whatis}
In recent years, we saw the formation of RSE institutes and advocacy groups, such as
the UK Software Sustainability Institute\footnote{\url{https://www.software.ac.uk}},
the US Research Software Sustainability Institute\footnote{\url{https://urssi.us}},
the European Virtual Institute for Research Software Excellence\footnote{\url{https://everse.software}},
and the EOSC Opportunity Area Expert Group 7 on Research Software%
\footnote{\url{https://eosc.eu/opportunity-area-exp/oa7-research-software}}.
We expect these initiatives to help increase the visibility and accelerate
the professionalisation%
\footnote{MM: do we want to decide for UK or US English? Or keep it mixed? Not being a native English speaker I don't know how it reads.}
of the RSE role.
In addition, a multitude of RSE communities self-organised at the national
and regional level to support more viable RSE career paths in academia.
We can expect RSEs to become more easily accessible,
possibly through centrally-funded long-term positions at research institutions.
The community of RSE's will get a significant boost in their acceptance if RSE Masters programmes
become available at various universities.

@@

\subsection{Impact of AI and automation}
\begin{whatis}{What do we want to have achieved}{impact-ai-automation}
RSEs are educated about AI models, and can help researchers reap their benefits, while
clearly knowing their limits.
\end{whatis}
AI is already transforming research processes by enabling rapid analysis of massive datasets, uncovering patterns and insights that would be impossible or prohibitively time-consuming for humans alone to find.
In research and development, AI accelerates scientific breakthroughs in fields like healthcare, climate science, and materials engineering by automating complex modeling and predictive tasks.
For software development, AI-powered tools automate code generation, bug detection, and testing, significantly increasing productivity and reducing human error.
Developers are shifting from manual coding to orchestrating AI-driven workflows, focusing more on high-level design, integration, and creative problem-solving while AI handles repetitive or routine tasks.
The integration of AI is democratizing access to advanced technologies and analytics, enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
As AI continues to evolve, it will further blur the lines between human and machine roles, making research and software development faster, more efficient, and more innovative than ever before.
With the general availability of AI the interaction with researchers will change at a fundamental level.
Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to it,
It can be foreseen, that AI drives this low-key interaction to a highly specialized level of knowledge,
since AI copilots are available 24/7.

@@

The ethics and values of the Research Software Engineer are grounded in their practice.
They commit to the health, safety and welfare of the public and act in the interest of society, their employer and their clients \cite{Goth2024}.

@@

\begin{whatis}{What do we want to have achieved}{education}
School curricula have embraced the digital transformation, and have included basics of computer interaction
in every class level.
In universities, the first generations of RSE-Master graduates find their way into research departments,
and accelerate the pace of science.
\end{whatis}

@@

\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
\begin{whatis}{What do we want to have achieved}{career-paths}
We would like to see the emergence of more dedicated personel, i.e. other RTPs next to the RSE.
\end{whatis}
RSEs can either be embedded in research groups (belonging to the mid-level university staff),
or be members of larger RSE units.\cite{Kempf2025-draft}
The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.
This, however, comes with additional funding challenges,
as long-term funding is more difficult to acquire in lower levels of an organization.
Third-party funding can mitigate this issue in the mid-term.
However, the participants considered central RSE groups more likely in the future,
due to the fact that there are already examples and activities from various universities in this direction
(examples include the universities of Jena, Heidelberg, Göttingen).
We expect funding in central groups to evolve from initially temporary positions to permanent positions,
as the groups themselves prove themselves valuable.
Embedded RSEs and central groups will need to network with each other,
and potentially request and coordinate support from external RSEs.
